Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-24 submission/manuscript.Rmd
add accent for naive

Distinguishing between the germline (cells that pass on their genetic material, e.g. sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis. In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na\"{i}ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters. Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.

@@

Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023]. Somatic repression of germline genes is typically established during the transition between na\"{i}ve and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation. In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Germline genes are then silenced long-term in EpiLCs by _de novo_ placement of DNA CpG methylation (DNAme)[@mochizukiRepressionGermlineGenes2021]. Although KDM5C's catalytic activity is not required to suppress _Dazl_ in ESCs, KDM5C may promote long-term germline gene silencing via H3K4me3 removal since H3K4me3 can impede DNAme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and DNAme is lost at select germline gene promoters in the hippocampus[@scandagliaLossKdm5cCauses2017]. Because KDM5C's role in germline gene repression has only been characterized in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during transition between ESCs and EpiLCs and if its catalytic activity is required for long-term silencing.

@@

We found misexpression of tissue-enriched genes, including germline genes, also occurs in _Kdm5c_-KO epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.

- _In vivo_, a subset of epiblast stem cells will reactivate germline gene expression and form primordial germ cells (PGCs) that will eventually become the mature germline. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX]

- Importantly, we did not observe any significant changes in the expression of EpiLC marker genes and cellular morphology, indicating _Kdm5c_-KO EpiLCs are not simply becoming PGCLCs. 
- This is further supported proper expression of _Otx2_ in _Kdm5c_-KO EpiLCs, an epiblast stem cell marker that is known to repress differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. This suggests aberrant germline gene expression is occuring ectopically in conjunction with typical developmental programs, in conjunction with _Kdm5c_-KO mice having overall unimpaired brain morphology and very few brain-specific DEGs.

- EpiLCs are an _in vitro_ model of the post-implantation embryo, when cells transition from na\"{i}ve to primed pluripotency.

@@

- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from na\"{i}ve to primed pluripotency.
	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
	- KDM5C could be important for DNAme is that KDM5C erases H3K4me3 which can impede CpGme.

@@

Interestingly, some of the ectopic testis transcripts identified in the _Kdm5c_-KO brain are typically expressed in germ cells[@scandagliaLossKdm5cCauses2017]. Unlike somatic cells, germ cells (e.g. sperm and eggs) undergo meiosis and pass on their genetic material to the next generation. The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline. Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters. KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017]. In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. In support of this, two independent screens in mouse embryonic stem cells (ESCs) recently identified KDM5C as a repressor of _Deleted in azoospermia like (Dazl)_, a key regulator of germline development. However, KDM5C's role in embryonic germline gene repression is currently unclear, given that Dazl is also expressed in ESCs and in the 2-cell stage and germline gene misexpression has yet to be globally characterized during _Kdm5c_-KO embryogenesis.

>>>>

Distinguishing between the germline (cells that pass on their genetic material, e.g. sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis. In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters. Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.

@@

Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023]. Somatic repression of germline genes is typically established during the transition between na&#239;ve and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation. In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Germline genes are then silenced long-term in EpiLCs by _de novo_ placement of DNA CpG methylation (DNAme)[@mochizukiRepressionGermlineGenes2021]. Although KDM5C's catalytic activity is not required to suppress _Dazl_ in ESCs, KDM5C may promote long-term germline gene silencing via H3K4me3 removal since H3K4me3 can impede DNAme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and DNAme is lost at select germline gene promoters in the hippocampus[@scandagliaLossKdm5cCauses2017]. Because KDM5C's role in germline gene repression has only been characterized in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during transition between ESCs and EpiLCs and if its catalytic activity is required for long-term silencing.

@@

We found this dysregulation in tissue identity likely begins during early emrbyogenesis, as germline genes and other tissue-enriched genes are aberrantly expressed in _Kdm5c_-KO epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX]. Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest they are progressing beyond EpiLC differentiation and becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs had proper expression of EpiLC marker genes for primed pluripotency and we observed no difference in cellular morphology during ESC to EpiLC differentiation. Furthermore, we saw no significant change in _Otx2_ expression, an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. This, in conjunction with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to germline identity.

@@

- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from na&#239;ve to primed pluripotency.
	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
	- KDM5C could be important for DNAme is that KDM5C erases H3K4me3 which can impede CpGme.

@@

Interestingly, some of the ectopic testis transcripts identified in the _Kdm5c_-KO brain are typically expressed in germ cells[@scandagliaLossKdm5cCauses2017]. Unlike somatic cells, germ cells (e.g. sperm and eggs) undergo meiosis and pass on their genetic material to the next generation. The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline. Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from na&#239;ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters. KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017]. In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. In support of this, two independent screens in mouse embryonic stem cells (ESCs) recently identified KDM5C as a repressor of _Deleted in azoospermia like (Dazl)_, a key regulator of germline development. However, KDM5C's role in embryonic germline gene repression is currently unclear, given that Dazl is also expressed in ESCs and in the 2-cell stage and germline gene misexpression has yet to be globally characterized during _Kdm5c_-KO embryogenesis.
